---
title: "Reproducible Research: Peer Assessment 1"
author: "Troy Rose <troyjrose@gmail.com>"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
````{r}
library(dplyr)
library(ggplot2)
activitydata<-read.csv("C:\\Users\\troy\\Desktop\\repdata-data-activity\\activity.csv")
# translate time interval to %H:%M
translate<-function(x){
    return(paste(sprintf("%02d", trunc(x / 100)),":", sprintf("%02d", x %% 100), sep=""))
}
# apply to interval to create "time" field
activitydata$time<-lapply(activitydata$interval, translate)

# create datetime field
activitydata$datetime<-as.POSIXct(strptime(paste(activitydata$date, " ", activitydata$time), format = "%Y-%m-%d %H:%M"))

```


## What is mean total number of steps taken per day?
```{r}
# group, summarise, removing missing values
activitydata.by.day<-group_by(activitydata, date)
summary.activitydata.by.day<-summarise(activitydata.by.day, total = sum(steps))
qplot(summary.activitydata.by.day$total, geom="histogram")
meansteps<-mean(summary.activitydata.by.day$total, na.rm=TRUE)
mediansteps<-median(summary.activitydata.by.day$total, na.rm=TRUE)

```
80
The mean steps per day is `r meansteps`.

The median steps per day is `r mediansteps`.

## What is the average daily activity pattern?
```{r,echoresult="false"}
# group, summarise, removing missing values
activitydata.by.interval<-group_by(activitydata,interval)
summary.activitydata.by.interval<-summarise(activitydata.by.interval, mean = mean(steps, na.rm=TRUE))
qplot(summary.activitydata.by.interval$interval,summary.activitydata.by.interval$mean,ylab="Average number of steps per interval", xlab="5 minute intervals over 24 hours") + geom_line(aes(y = summary.activitydata.by.interval$mean), size = .5, alpha = 1)
maxmeansteps<-max(summary.activitydata.by.interval$mean, na.rm=TRUE)
intmaxmeansteps<-filter(summary.activitydata.by.interval, mean==maxmeansteps)$interval
```

The maximun of the mean total steps per day is `r maxmeansteps`.

The 5-minute interval that, on average, contains the maximum number of steps  is `r intmaxmeansteps`.

## Inputing missing values
```{r}
totalnas<-length(activitydata$steps[is.na(activitydata$steps)])

fillnas<-function(x) {
        if(is.na(x$steps)) {
                trunc(filter(summary.activitydata.by.interval, interval==x$interval)$mean)
        } else {
                return(x$steps)
        }
}
activitydata$stepsfilled<-apply(activitydata, c(1),fillnas)

newactivitydata.by.day<-group_by(activitydata, date)
summary.newactivitydata.by.day<-summarise(newactivitydata.by.day, total = sum(steps))
#hist(summary.newactivitydata.by.day$total,  main = "Histogram of steps per day")
qplot(summary.newactivitydata.by.day$total, geom="histogram")
newmeansteps<-mean(summary.newactivitydata.by.day$total, na.rm=TRUE)
newmediansteps<-median(summary.newactivitydata.by.day$total, na.rm=TRUE)


```
Number of NAs for steps is `r totalnas`.

My strategy for sumplementing missing values for steps is:

1. For all missing values:

        1. Replace the missing value with the average for that interval across all days truncated to a whole integer.
        
The new steps filled in are stored in activitydata$stepsfilled.

The mean steps per day is `r newmeansteps`.

The median steps per day is `r newmediansteps`.

It does not appear that filling in the missing values of steps made very much difference to the overall mean and median steps per day.

## Are there differences in activity patterns between weekdays and weekends?

```{r}

# create new factor with two levels for days of week in weekend and weekday
activitydata$dayofweek<-factor(c("weekend", "weekday", NA))

# fill in values for weekends and weekdays
activitydata[!(weekdays(as.Date(activitydata$date)) %in% c('Saturday','Sunday')),]$dayofweek<-as.factor("weekday")
activitydata[(weekdays(as.Date(activitydata$date)) %in% c('Saturday','Sunday')),]$dayofweek<-as.factor("weekend")

# group by day of week and interval
activitydata.by.dayofweek<-group_by(activitydata, dayofweek, interval)

# summarise data getting average of steps per day for each weekend and weekday
summary.activitydata.by.dayofweek<-summarise(activitydata.by.dayofweek, mean = mean(stepsfilled))

# plot data
qplot(interval, mean, data=summary.activitydata.by.dayofweek, ylab="Average number of steps per interval", xlab="5 minute intervals over 24 hours",facets = dayofweek ~ . ) + geom_line(aes(y = summary.activitydata.by.dayofweek$mean), size = .5, alpha = 1)
```

Yes, it appears that on weekends, there is less early morning activity, with peak activity occuring at around 8:00am, with very little activity prior to 8:00am (it appears that people like to sleep in on weekends).

This is different to the weekdays where morning activity starts at around 5:00am and peaks at around 9:00am. 

End of analysis.